Closure-Free Functional Programming in a Two-Level Type Theory

### 2024-06-07 newpaper/paper.tex
after appendix finish

\newcommand{\mit}[1]{{\mathsf{#1}}}
\newcommand{\msf}[1]{{\mathsf{#1}}}
\newcommand{\mbf}[1]{{\mathbf{#1}}}
\newcommand{\mbb}[1]{\mathbb{#1}}
\newcommand{\rel}{^{\approx}}
\newcommand{\U}{\mathsf{U}}
\newcommand{\p}{\mathsf{p}}
\newcommand{\q}{\mathsf{q}}
\newcommand{\code}{\mathsf{code}}
\newcommand{\bs}[1]{\boldsymbol{#1}}
\newcommand{\wh}[1]{\widehat{#1}}
\newcommand{\mdo}{\mbf{do}\,}
\newcommand{\ind}{\hspace{1em}}
\newcommand{\bif}{\mbf{if}\,}
\newcommand{\bthen}{\mbf{then}\,}
\newcommand{\belse}{\mbf{else}\,}
\newcommand{\return}{\mbf{return}\,}
\newcommand{\pure}{\mbf{pure}\,}
\newcommand{\lam}{\lambda\,}
\newcommand{\data}{\mbf{data}\,}
\newcommand{\where}{\mbf{where}}
\newcommand{\M}{\msf{M}}
\newcommand{\letrec}{\mbf{letrec}\,}
\newcommand{\of}{\mbf{of}\,}
\newcommand{\go}{\mit{go}}
\newcommand{\add}{\mit{add}}
\newcommand{\letdef}{\mbf{let\,}}
\newcommand{\map}{\mit{map}}
\newcommand{\emptycon}{\scaleobj{.75}\bullet}
\newcommand{\Tyo}{\msf{Ty}_{\mbbo}}
\newcommand{\Tmo}{\msf{Tm}_{\mbbo}}
\newcommand{\whset}{\wh{\Set}}
\newcommand{\ev}{\mbb{E}}
\newcommand{\re}{\mbb{R}}
\newcommand{\welim}{\vW{-}\msf{elim}}

@@

\appendices
\section{Staged Semantics of CFTT}\label{appendix:A}

@@

Next, we give a logical relation interpretation for $\vW$. We define an
inductive relation in $\whset$ as follows.
\begin{alignat*}{3}
  & \vW\rel &&: \Gamma\rel\,\gamma\,\gamma' \to \ev\,(\vW\,A\,B)\,\gamma \to \re\,(\vW\,A\,B)\,\gamma' \to \whset\\
  & \vsup\rel &&: \hspace{0.8em}\{\alpha : \ev\,A\,\gamma\}\{\alpha' : \re\,A\,\gamma'\}(\alpha\rel : A\rel \gamma\rel\,\alpha\,\alpha')\\
  &           &&\to \{f \hspace{0.5em}: \ev\,B\,(\gamma,\,\alpha) \to \ev\,(\vW\,A\,B)\,\gamma\}\\
  &           &&\to \{f' \hspace{0.2em}: \re\,(B \Rightarrow (\vW\,A\,B)[\p])\,(\gamma',\,\alpha')\}\\
  &           &&\to (f\rel : (B \Rightarrow (\vW\,A\,B)[\p])\rel\,(\gamma\rel,\,\alpha\rel)\,f\,f')\\
  &           &&\to \vW\rel\,\gamma\rel\,(\wh{\vsup}\,\alpha\,f)\,(\re_\vsup\,\alpha'\,f')
\end{alignat*}
Here, $(B \Rightarrow (\vW\,A\,B)[\p])\rel$ is the relational interpretation of the function type, which unfolds to
pointwise preservation of relations. In short, $\vW\rel$ expresses that $\wh{\vsup}\,\alpha\,f$ is related $\re_\vsup\,\alpha'\,f'$
when the arguments (inductively) related. We can immediately interpret $\vW$ and $\vsup$:
\begin{alignat*}{3}
  & (\vW\,A\,B)\rel\,\gamma\rel\,t\,t' &&:= \vW\rel\,\gamma\rel\,t\,t'\\
  & (\vsup\,t\,u)\rel\,\gamma\rel      &&:= \vsup\rel\,(t\rel\,\gamma\rel)\,(u\rel\,\gamma\rel)
\end{alignat*}
It only remains to interpret $\vW$-elimination and its $\beta$-rule. We need to define
\[ (\vW{-}\msf{elim}\,P\,s\,w)\rel\,\gamma\rel : P\rel\,(\gamma\rel,\,w\rel\,\gamma\rel)\,(\ev\,(\welim\,P\,s\,w)\,\gamma)\,(\re\,(\welim\,P\,s\,w)\,\gamma') \]
We do this by induction on $w\rel\,\gamma\rel :
\vW\rel\,\gamma\rel\,(\ev\,w\,\gamma)\,(\re\,w\,\gamma')$. At this point it
becomes rather tedious to compute and unfold the involved types, so we only give
a compact definition here together with some informal explanation:
\begin{alignat*}{3}
  &(\vW{-}\msf{elim}\,P\,s\,w)\rel\,\gamma\rel :=\\
  &\hspace{4em}\vW\rel{-}\msf{elim}\,(\lambda\,w\rel.\,\,P\rel\,(\gamma\rel,\,w\rel)\,(\ev\,(\welim\,P\,s\,w)\,\gamma)\,(\re\,(\welim\,P\,s\,w)\,\gamma'))\\
  &\hspace{8.1em}(\lambda\,a\rel\,f\rel\,\mathit{hyp}.\,\,s\rel\,(\gamma\rel,\,a\rel,\,f\rel,\,\mathit{hyp}))\\
  &\hspace{8.1em}(w\rel\,\gamma\rel)
\end{alignat*}
$\vW\rel{-}\msf{elim}$ is the eliminator for $\vW\rel$. Its first argument is
the induction motive, which matches the goal type. The second argument is the
method for $\vsup\rel$; its $a\rel$ and $f\rel$ inputs are the relatedness
witnesses stored in a $\vsup\rel$, while $\mathit{hyp}$ is induction hypothesis,
witnessing the inductive motive for every output of $f\rel$. In the body here,
we need to witness the motive in the case where $w$ is a $\vsup$, which causes
both $\ev\,(\welim\,P\,s\,w)\,\gamma)$ and $\re\,(\welim\,P\,s\,w)\,\gamma')$ to
compute further to applications of $\ev\,s\,\gamma$ and $\re\,s\,\gamma'$
respectively. Hence, we use $s\rel$ to relate the two sides. The
$\vW{\beta}$-rule also evidently holds with this definition.

@@

>>>>

\newcommand{\mit}[1]{{\mathsf{#1}}}
\newcommand{\msf}[1]{{\mathsf{#1}}}
\newcommand{\mbf}[1]{{\mathbf{#1}}}
\newcommand{\mbb}[1]{\mathbb{#1}}
\newcommand{\rel}{^{\approx}}
\newcommand{\U}{\mathsf{U}}
\newcommand{\Id}{\msf{Id}}
\newcommand{\p}{\mathsf{p}}
\newcommand{\q}{\mathsf{q}}
\newcommand{\code}{\mathsf{code}}
\newcommand{\bs}[1]{\boldsymbol{#1}}
\newcommand{\wh}[1]{\widehat{#1}}
\newcommand{\mdo}{\mbf{do}\,}
\newcommand{\ind}{\hspace{1em}}
\newcommand{\bif}{\mbf{if}\,}
\newcommand{\bthen}{\mbf{then}\,}
\newcommand{\belse}{\mbf{else}\,}
\newcommand{\return}{\mbf{return}\,}
\newcommand{\pure}{\mbf{pure}\,}
\newcommand{\lam}{\lambda\,}
\newcommand{\data}{\mbf{data}\,}
\newcommand{\where}{\mbf{where}}
\newcommand{\M}{\msf{M}}
\newcommand{\letrec}{\mbf{letrec}\,}
\newcommand{\of}{\mbf{of}\,}
\newcommand{\go}{\mit{go}}
\newcommand{\add}{\mit{add}}
\newcommand{\letdef}{\mbf{let\,}}
\newcommand{\map}{\mit{map}}
\newcommand{\emptycon}{\scaleobj{.75}\bullet}
\newcommand{\Tyo}{\msf{Ty}_{\mbbo}}
\newcommand{\Tmo}{\msf{Tm}_{\mbbo}}
\newcommand{\Cono}{\msf{Con}_{\mbbo}}
\newcommand{\Subo}{\msf{Sub}_{\mbbo}}
\newcommand{\whset}{\wh{\Set}}
\newcommand{\ev}{\mbb{E}}
\newcommand{\re}{\mbb{R}}
\newcommand{\welim}{\vW{-}\msf{elim}}

@@

\appendices
\section{Staging in CFTT}\label{appendix:A}

@@

Next, we give a logical relation interpretation for $\vW$. We define an
inductive relation in $\whset$ as follows.
\begin{alignat*}{3}
  & \vW\rel &&: \Gamma\rel\,\gamma\,\gamma' \to \ev\,(\vW\,A\,B)\,\gamma \to \re\,(\vW\,A\,B)\,\gamma' \to \whset\\
  & \vsup\rel &&: \hspace{0.8em}\{\alpha : \ev\,A\,\gamma\}\{\alpha' : \re\,A\,\gamma'\}(\alpha\rel : A\rel \gamma\rel\,\alpha\,\alpha')\\
  &           &&\to \{f \hspace{0.5em}: \ev\,B\,(\gamma,\,\alpha) \to \ev\,(\vW\,A\,B)\,\gamma\}\\
  &           &&\to \{f' \hspace{0.2em}: \re\,(B \Rightarrow (\vW\,A\,B)[\p])\,(\gamma',\,\alpha')\}\\
  &           &&\to (f\rel : (B \Rightarrow (\vW\,A\,B)[\p])\rel\,(\gamma\rel,\,\alpha\rel)\,f\,f')\\
  &           &&\to \vW\rel\,\gamma\rel\,(\wh{\vsup}\,\alpha\,f)\,(\re_\vsup\,\alpha'\,f')
\end{alignat*}
Here, $(B \Rightarrow (\vW\,A\,B)[\p])\rel$ is the relational interpretation of the function type, which unfolds to
pointwise preservation of relations. In short, $\vW\rel$ expresses that $\wh{\vsup}\,\alpha\,f$ is related $\re_\vsup\,\alpha'\,f'$
when the arguments (inductively) related. We can immediately interpret $\vW$ and $\vsup$:
\begin{alignat*}{3}
  & (\vW\,A\,B)\rel\,\gamma\rel\,t\,t' &&:= \vW\rel\,\gamma\rel\,t\,t'\\
  & (\vsup\,t\,u)\rel\,\gamma\rel      &&:= \vsup\rel\,(t\rel\,\gamma\rel)\,(u\rel\,\gamma\rel)
\end{alignat*}
It only remains to interpret $\vW$-elimination and its $\beta$-rule. We need to define
\[ (\vW{-}\msf{elim}\,P\,s\,w)\rel\,\gamma\rel : P\rel\,(\gamma\rel,\,w\rel\,\gamma\rel)\,(\ev\,(\welim\,P\,s\,w)\,\gamma)\,(\re\,(\welim\,P\,s\,w)\,\gamma') \]
We do this by induction on $w\rel\,\gamma\rel :
\vW\rel\,\gamma\rel\,(\ev\,w\,\gamma)\,(\re\,w\,\gamma')$. At this point it
becomes rather tedious to compute and unfold the involved types, so we only give
a compact definition here together with some informal explanation:
\begin{alignat*}{3}
  &(\vW{-}\msf{elim}\,P\,s\,w)\rel\,\gamma\rel :=\\
  &\hspace{4em}\vW\rel{-}\msf{elim}\,(\lambda\,w\rel.\,\,P\rel\,(\gamma\rel,\,w\rel)\,(\ev\,(\welim\,P\,s\,w)\,\gamma)\,(\re\,(\welim\,P\,s\,w)\,\gamma'))\\
  &\hspace{8.1em}(\lambda\,a\rel\,f\rel\,\mathit{hyp}.\,\,s\rel\,(\gamma\rel,\,a\rel,\,f\rel,\,\mathit{hyp}))\\
  &\hspace{8.1em}(w\rel\,\gamma\rel)
\end{alignat*}
$\vW\rel{-}\msf{elim}$ is the eliminator for $\vW\rel$. Its first argument is
the induction motive, which matches the goal type. The second argument is the
method for $\vsup\rel$; its $a\rel$ and $f\rel$ inputs are the relatedness
witnesses stored in a $\vsup\rel$, while $\mathit{hyp}$ is induction hypothesis,
witnessing the inductive motive for every output of $f\rel$. In the body here,
we need to witness the motive in the case where $w$ is a $\vsup$, which causes
both $\ev\,(\welim\,P\,s\,w)\,\gamma)$ and $\re\,(\welim\,P\,s\,w)\,\gamma')$ to
compute further to applications of $\ev\,s\,\gamma$ and $\re\,s\,\gamma'$
respectively. Hence, we use $s\rel$ to relate the two sides. This definition
also respects $\vW\beta$, by the $\beta$ rule for $\vW\rel$.

@@

In this section we assume an \emph{extensional} identity type in CFTT.  As we
mentioned in Section \ref{TODO}, we can show soundness of staging with
extensional identity and certain \emph{postulated} identities, but we cannot do
the same with intensional identity. Hence, we formulate unstaging as consisting
of two steps: first, a translation of intensional identity to extensional (which
amounts to proof erasure), followed by evaluation in the presheaf model.

Extensional identity in CFTT type is interpreted as the standard extensional
identity in presheaves, i.e.\ we have $\ev\,(\msf{Id}\,A\,t\,u)\,\gamma :=
\ev\,t\,\gamma = \ev\,u\,\gamma$. This supports $\msf{refl}$, transport,
uniqueness of identity and equality reflection; we omit the fairly trivial
definitions here. We choose the relational interpretation to be trivial as well,
i.e.\ $(\msf{Id}\,t\,u)\rel\,\gamma\rel\,p\,p' := \top$.

This is because both $\ev\,(\Id\,t\,u)\,\gamma$ and $\re\,(\Id\,t\,u)\,\gamma'$
are mere propositions --- the latter because of equality reflection in the
syntax.  In particular, if we have a witness of $\re\,(\Id\,t\,u)\,\gamma'$, we
also have $\re\,t\,\gamma' = \re\,u\,\gamma'$ by equality reflection. Thus, the
rest of the relational interpretation of $\Id$ is also trivial. Another way to
think about this is that soundness expresses a form of canonicity, but if we
have equality reflection then syntactic identities are always canonical.

\subsection{Generativity}

The generativity axiom from Section \ref{TODO} has the following type, in internal
CFTT syntax:
\[ (A : \U_\msf{P})(f : \Elprod\,A \to \USOP)(x\,y : \Elprod\,A) \to f\,x = f\,y  \]
In the presheaf model, we have the following:
\begin{itemize}
\item $\wh{\Uprod}$ is the set (constant presheaf) of lists of object-theoretic types.
\item $\wh{\USOP}$ is the set of lists of lists of object-theoretic types.
\item $\wh{\Elprod\,A}$ is the presheaf such that $\wh{\Elprod\,A}\,\Gamma$ is
  the set of products of object-theoretic terms in context $\Gamma$, with types
  of components taken from $A$.
\end{itemize}
Unfolding the type of generativity in the presheaf model, it suffices to show
that assuming
\begin{itemize}
\item $A : \wh{\Uprod}$, $\Gamma : \Cono$, $x,\,y : \wh{\Elprod\,A}\,\Gamma$
\item and $f : (\Delta : \Cono) \to \Subo\,\Delta\,\Gamma \to \wh{\Elprod\,A}\,\Delta \to \wh{\USOP}$ such that $f$ is natural with respect to object-theoretic substitution,
\end{itemize}
we have that $f\,\Gamma\,\id\,x = f\,\Gamma\,\id\,y$.

Now, $\wh{\Elprod\,A}$ is a representable presheaf since it is a finite product
of object-theoretic terms. Also, since $f$ is a natural transformation from the
representable presheaf $\Subo\blank\Gamma \times \wh{\Elprod\,A}$ to the
constant presheaf $\wh{\USOP}$, the Yoneda lemma implies that $f$ is constant and
hence the generativity axiom is validated.

Since the logical relation interpretation of identity is trivial, it is also trivial
for the generativity axiom.

\subsection{Summary}

By the presheaf and logical relations interpretation of CFTT as shown in this appendix,
we obtain \emph{unstaging} and \emph{soundness} of unstaging, following \cite{staged2ltt}.

\emph{Completeness} of unstaging is shown by straightforward induction on the
syntax of the object theory, the same way as in \cite{staged2ltt}, and we omit
it here.
